The Talking Heads experiment

### 2015-04-29 chap7/chap7.tex
format relevant papers section
part 2 of martin haspelmath's comments

Wittgenstein proposed to study the use of language in terms of 
games. This notion captures that the most basic form of
language involves a social interaction between individuals
within a specific setting, which acts as a context restricting
the possible meanings. The interaction is played out 
along a conventionalised set of rules, as is chess or any
other game. The notion of a game captures
that a language interaction has a purpose, for example,
to identify an object in a context, to gather 
information, to transmit emotion, to try and
invoke an action by another person, etc. Words get \enlargethispage{1\baselineskip}
their meaning as part of a language game. Language is a tool 
fully integrated in the rest of human social activity.\footnote{See in particular \cite{Wittgenstein:1953}.}

@@

For the interpretation of an utterance, the main 
flow of information goes in the other direction. The
lexical layer sends a variety of hypotheses to 
the conceptual layer and the conceptual layer makes use of data 
from the perceptual layer to see which conceptualisation
yields a referent. Several solutions are considered 
by the conceptual layer because words are typically 
ambiguous and so the perceptual 
layer must produce the appropriate segmentations and 
sensory characteristics to test each solution.
Even if a sensory channel was not considered to be 
very salient by the hearer, it may have been 
used by the speaker in his conceptualisation of the 
scene and hence the hearer's perceptual 
processes must actively try to seek in the 
image the information required to see whether it 
is applicable to the specific
context. The interpretation of an utterance
thus strongly takes the real world context into account. 
The utterance literally influences the way the 
hearer sees the world. The two-way flow (from perception
to conceptualisation and language and from language
and conceptualisation to perception) resolves one of the 
paradoxes of meaning discussed in Chapter 4. The 
clear picture of reality that we consciously experience
results from a dynamical process in which local constraints 
keep propagating until a globally coherent solution
emerges.\footnote{Dynamical systems which enter into an attractor
based on a similar relaxation process have been 
widely studied, particularly in physics. One of the 
prototypical examples is a spin glass whose magnetic
states keep switching until a globally coherent
state is reached.}

@@

\begin{figure}
\begin{center}
\includegraphics[width=0.8\columnwidth]{chap7/figs/plate-10.pdf}
\end{center}
\caption{Three examples of segmented images. The 
topic is indicated by a dashed bounding box in the 
image of the speaker. Segments which are too small 
are ignored. The topics have all been conceptualised
as being `to the right' and so the same word 
`gofubo' has been used to refer to them. }
\label{fig:plate-10}
\end{figure}

@@

The left image is that of the hearer {\bfshape a1}, the right one
that of the speaker, {\bfshape a2}.
{\bfshape a2} had already invented the word `gofubo' for 
[\textsc{hpos} 0.5–1.0] (to the right) in an earlier game, 
but the word was not acquired in that game
by {\bfshape a1} because he still missed the 
appropriate distinction. Meanwhile the \textsc{hpos} category 
is available to {\bfshape a1} due to an
expansion of his discrimination networks
and so he can store the word: 
\begin{verbatim}
Game 3 
  a2 is the speaker. a1 is the hearer. 
  a2 segments the context into 2 objects: 
       object-0 object-1
  a2 chooses object-0 as the topic 
  a2 categorises the topic as [\textsc{hpos} 0.5–1.0]
  a2 says: gofubo
  a1 does not know gofubo
  a1 says: gofubo?
  a2 points to object-0
  a1 categorises the topic as [\textsc{hpos} 0.5–1.0]
  a1 stores gofubo as [\textsc{hpos} 0.5–1.0]
\end{verbatim}
This game proceeds essentially like in the computer
simulations studied before. The major difference is
that now real images have been used, the objects
considered are the outcome of segmentation processes,
the sensory characteristics have been derived from the 
image itself, and the pointing has been done by 
physically moving the cameras.

@@

Immediately the agents apply this word to very 
different scenes, such as the ones in \figref{fig:plate-10} (middle and bottom). The middle picture
shows on the left the segmented scene
from the speaker in game 5, and on the right the one
from the hearer. In game 5, the figures are blue
rectangles. They are much further apart than 
the two circles in game 3. Nevertheless, 
after scaling, they are categorised and
conceptualised similarly and therefore
the same word could be used effectively. The concept 
of [\textsc{hpos} 0.5–1.0] and hence the word `gofubo' 
is general {\itshape from the very start}. 
The agents do not need to see many examples because they do 
not use inductive generalisation. Instead, 
the \textsc{hpos} category is constructed in a top-down fashion 
as soon as the \textsc{hpos} channel has been salient
and is immediately available for use in the 
discrimination game and hence in verbalising 
the scene. This explains why the word-meaning acquisition
process observed in the experiments goes so 
amazingly fast.

The bottom of \figref{fig:plate-10}
shows yet another scene where the word `gofubo' 
was used with success. The topic is the rightmost
shape in the scene and so [\textsc{hpos} 0.5–1.0] is once more 
distinctive. To an outside observer it may look 
like the agents have performed a gigantic inductive
leap, but this is not the case at all. The agents
do not try to abstract the commonalities from 
different examples but construct distinctions in a top 
down fashion and try to apply them to the perceived image.

@@

There is already a word (`gofubo') which has
a score of 1.0! After 100 games, the lexicon has become more 
solid and now looks as in \tabref{tab:goubo}.

@@

\begin{table}
\begin{center}
\begin{tabular}{lll}
\lsptoprule
{\itshape channel}& {\itshape obj-0} & {\itshape obj-1}\\ \midrule
\textsc{hpos} & 0.08 & 0.09\\ 
\textsc{vpos} & 0.21 & 0.10\\ 
\textsc{height} & 0.10 & 0.15\\ 
\textsc{width} & 0.42 & 0.11\\ 
\textsc{area} & 0.34 & 0.16\\ 
\textsc{r} & 0.33 & 0.31\\ 
\textsc{g} & 0.68 & 0.65\\ 
\textsc{b} & 0.52 & 0.50\\ 
\lspbottomrule
\end{tabular}
\caption{\label{tab:game126}Sensory data for game 126.}
\end{center}
\end{table}
Clearly width is the most salient channel for the speaker. 
It is therefore chosen and because the 
agents have already grown distinctions on this channel, 
discrimination succeeds and a new word can be
constructed and stored. 
\begin{verbatim}
Game 126 
  a2 is the speaker. a1 is the hearer. 
  a2 segments the context into 2 objects: 
       object-0 object-1
  a2 chooses object-0 as the topic 
  a2 categorises the topic as [\textsc{width} 0.5–1.0]
  a2 creates a new word: vaviwumu
  a2 says: vaviwumu
  a1 does not know vaviwumu
  a1 says: vaviwumu?
  a2 points to object-0
  a1 categorises the topic as [\textsc{width} 0.5–1.0]
  a1 stores vaviwumu as [\textsc{width} 0.5–1.0]
\end{verbatim}
\figref{fig:plate-13} (bottom) contains another example
of segmented images which have stimulated conceptual
growth and hence an expansion of the lexicon. 
In this case, categorisations on the \textsc{height} channel are
relevant and a word developed for tall.

@@

\begin{figure}
\begin{center}
\includegraphics[width=0.8\columnwidth]{chap7/figs/plate-15-vpos.pdf}
\end{center}
\caption{The image segments from the top game require 
distinctions on the \textsc{vpos} channel and 
caused the creation of a word for `upper'. The image 
segments in the middle lead to the use of refined
distinctions on the \textsc{hpos} channel, so as to 
identify the middle square. The topic in the 
image segments at the bottom was done with a 
conjunction of categories: blue and tall.}
\label{fig:plate-15}
\end{figure}

@@

\begin{table}
\begin{center}
\begin{tabular}{lllll}
\lsptoprule
{\itshape Word}&{\itshape Meaning}&{\itshape Translation} & {\bfshape a1}&{\bfshape a2} \\ \midrule
wawosido & [\textsc{hpos} 0.0–0.5] &left&0.9&0.4\\ 
gixepo & [\textsc{hpos} 0.0–0.25] & very left&0.4&0.4\\ 
wonuxa & [\textsc{hpos} 0.0–0.25] & very left&0.10&0.0\\ 
meluri & [\textsc{hpos} 0.25–0.5] &medium left&0.4&0.4\\ 
gofubo & [\textsc{hpos} 0.5–1.0]& right&1.0&1.0\\ 
vokomutu & [\textsc{hpos} 0.5–0.75] &medium right&0.4&0.4\\ 
buwonipo & [\textsc{hpos} 0.75–0.875] &strongly right&0.1&0.0\\ 
wiwigapo & [\textsc{vpos} 0.0–0.5] &down&0.5&0.4\\ 
putuwenu & [\textsc{vpos} 0.5–1.0]&up & 0.5&0.5\\ 
vaviwumu & [\textsc{width} 0.5–1.0]&wide & 0.5&0.5\\ 
pesidumu & [\textsc{area} 0.0–0.5]&small& 0.2&0.2\\ 
fozumoba & [\textsc{area} 0.5–1.0]&large & 1.0&1.0\\ 
wefoto & [\textsc{r} 0.0–0.5]& low redness &0.2&0.1\\ 
togene & [\textsc{r} 0.5–1.0]& high redness &0.9&1.0\\ 
fumudanu & [\textsc{g} 0.0–0.5]& low greenness &0.7&0.9\\ 
puxedu & [\textsc{g} 0.5–1.0]& high greenness &0.9&0.9\\ 
\lspbottomrule
\end{tabular}
\caption{\label{tab:upper}Population lexicon after 200 games.}
\end{center}
\end{table}
Note that new words have come into the lexicon 
for up (`putuwenu'), down (`wiwigapo'), and 
wide (`vaviwumu'). The discrimination trees of 
the \textsc{width} and \textsc{height} channel have started to 
expand (\figref{discri200} bottom).

@@

\begin{table}
\begin{center}
\begin{tabular}{llll}
\lsptoprule
{\itshape channel}& {\itshape obj-0} & {\itshape obj-1} & {\itshape obj-2}\\ 
\textsc{hpos} & 0.03 & 0.15 & 0.33\\ 
\textsc{vpos} & 0.29 & 0.29 & 0.01\\ 
\textsc{height} & 0.37 & 0.39 & 0.07\\ 
\textsc{width} & 0.10 & 0.08 & 0.27\\ 
\textsc{area} & 0.29 & 0.22 & 0.16\\ 
R & 0.95 & 0.97 & 0.97 \\ 
G & 0.33 & 0.40 & 0.93\\ 
B & 0.35 & 0.44 & 0.26\\ 
\lspbottomrule
\end{tabular}
\caption{\label{tab:game128}Sensory data for game 128.}
\end{center}
\end{table}
Context-scaling further
amplifies this difference which shows that it is not 
always beneficial to do so. {\bfshape a1} uses
the horizontal axis, creating a new word `bagaxe' 
meaning `very left' [\textsc{hpos} 0.0–0.25] whereas {\bfshape a2}, for
whom the area is the most salient characteristic, 
associates this word with [\textsc{area} 0.5–1.0] (large). 
This example shows that grounding introduces 
additional risks for the introduction of semantic
incoherence in a group's lexicon.

Due to perceptual anomalies, a game may fail even though 
both agents already have the same meaning for the 
same word. This happens when this meaning yields different 
objects (or no objects at all) for the speaker and the hearer. 
As a consequence, the hearer adopts another meaning for 
the word which starts to compete with the one he already had. 
This happened in the following game (game 127)
based on the segmented images shown in \figref{fig:plate-16} (bottom). The speaker (left image) has identified
only two objects, but the hearer three. The 
speaker (left image) uses green as distinguishing 
characteristic which is indeed appropriate. but 
because the hearer's third object (the top rectangle) is 
also green, this distinction fails for him.  
Even though the hearer had already a well established
meaning for `puxedu', he associates a new meaning 
based on the \textsc{width}-channel with this word because this 
channel is now the most salient. 
\begin{verbatim}
Game 127
  a1 is the speaker. a2 is the hearer. 
  a1 segments the context into 2 objects: 
       s-object-0 s-object-1
  a1 chooses object-0 as the topic 
  a1 considers as salient G R AREA HPOS 
  a1 categorises the topic as 
       [HPOS 0.0–0.5] [HPOS 0.0–0.125]
       [AREA 0.0–0.5] [R 0.0–0.5] [G 0.5–1.0]
  a1 has the words
       puxedu for [G 0.5–1.0] (1.0)
       wawosido for [HPOS 0.0–0.5] (0.7)
       pesidimu for [AREA 0.0–0.5] (0.2)
       wefoto for [R 0.0–0.5] (0.2)
       buwonipo for [G 0.5–1.0] (0.0)
  a1 says: puxedu
  a2 segments the context into 3 objects: 
       h-object-0 h-object-1 h-object-2
  a2 interprets puxedu as
       [G 0.5–1.0] (0.20)
  a2 identifies h-object-0 h-object-1
  a2 says: puxedu?
  a1 points to s-object-0
  a2 categorises the topic as [WIDTH 0.5–1.0]
  a2 stores puxedu as [WIDTH 0.5–1.0]
\end{verbatim}

@@

The partial landscape in \figref{RMF1} (taken from
an actual experiment) contains an example where the agents use
two possible meanings for conceptualising {\itshape situation2}, namely
[\textsc{gray} 0.0–0.25] (very light) and [\textsc{hpos} 0.5–0.75]
(medium right ). The words `katapu' and `tisame' lexicalise
[\textsc{gray} 0.0–0.25] and `wobo' and `tisame' [\textsc{hpos} 0.5–0.75]. 
Each meaning has therefore two synonyms and `tisame' is 
ambiguous; it can mean both [\textsc{gray} 0.0–0.25]
and [\textsc{hpos} 0.5–0.75].
\enlargethispage{1em}
Three words
are used to communicate the situation: `katapu', `tisame', 
and `wobo'. Such a structure is typical for grounded
lexicon evolution and complexity rapidly increases when the
same meanings are used to communicate about other
situations (which is obviously very common and indeed desirable).

@@

The RF-diagram\is{RF-diagram} in \figref{RF-diagram} shows the 
frequency with which certain words were used to communicate 
a particular situation. We see that in
the beginning the word `desepu' has been dominant, then 
there is a period of turbulence in which different 
words compete, but after a while a new word `va' wins 
the competition and becomes the dominant way to communicate 
about this situation. \figref{FM-diagram} shows another\is{FM-diagram}
competition diagram plotting the evolution of the FM-relation, for 
the word form `va', in other words the frequencies of 
all the meanings that co-occurred with the word `va'. 
We see an early peak when `va' was used in 70 \% of the 
games with the meaning [\textsc{b} 0.3125–0.375], i.e. a particular
shade of blue. Then there is a struggle during which additional
distinctions (on the \textsc{red} and \textsc{vpos}-channels) are competing
for the dominant meaning of `va'. 
[\textsc{r} 0.0–0.125] finally becomes dominant.

\begin{figure}[htbp]
  \centerline{\includegraphics[width=.80\textwidth]{chap7/figs/fm.pdf}}
\caption{\label{FM-diagram}This FM-diagram shows
the frequencies of each form-meaning pair with 
the form equal to `va' in a series of 5000 games.
A disambiguating situation occurs in game 3000
causing the loss of one meaning of `va'.}
\end{figure}

These competition diagrams are an important tool to try 
and make sense of the ontological and lexical evolution
taking place in evolving groups of agents as they are
playing their language games. Typically we pick
a dominating word, for example the word `va' in 
\figref{RF-diagram}, and try to understand
why it has become dominant. 
The FM-diagram for `va' (\figref{FM-diagram}) explains
part of the story. Three stable meanings for `va' have emerged
at  around 1000 games: 
[\textsc{r} 0.0–0.125], [\textsc{b} 0.3125–0.375], and [\textsc{vpos} 0.25–0.5]. 
They are all equally adequate for distinguishing the object
`va' designates, and there are no situations yet
that would have forced the disambiguation of `va'. 
In game 3000, the environment (which is continuously 
changing in this experiment) produces a scene in which
a category which was distinctive for the object designated by 
`va' is no longer distinctive. The lexicon adapts 
immediately. Around game 3000 the \textsc{vpos}-based meaning disappears, 
and the distinction based on \textsc{red} shoots up and becomes dominant.

@@

\subsection{The story of `fepi'}

\begin{figure}[htbp]
  \centerline{\includegraphics[width=.80\textwidth]{chap7/figs/FR-FEPI.pdf}}
\caption{\label{fr-fepi}FR-diagram showing the frequencies
of the objects referred to by the word
`fepi'. `fepi' is consistently used for \emph{O3} as well as
\emph{O5}.}
\end{figure}
Only looking at macroscopic measures like communicative
success hides away the
interesting rich lexical dynamics that unfolds
in the population. Let me examine just one word, `fepi'. 
By looking at the FR-diagram, we see that this word is 
used consistently for identifying two objects
(\emph{O3} and \emph{O5}) in a certain set of contexts
(\figref{fr-fepi}).

\begin{figure}[htbp]
  \centerline{\includegraphics[width=.80\textwidth]{chap7/figs/FM-FEPI.pdf}}
\caption{\label{fm-fepi}FM-diagram showing the different
meanings of `fepi'. One meaning [\textsc{g} 0.25–0.5] dominates.}
\end{figure}
We can see the meanings of `fepi', by 
inspecting its FM-diagram (\figref{fm-fepi}). 
The dominant meaning of `fepi' is 
a particular shade of green [\textsc{g} 0.25–0.5]. There are 
some other competing meanings (including [\textsc{gray} 0.25–0.3125])
but most of them are hardly ever used. We observe clearly 
the tendency for ambiguity to get damped.

What about synonymy? Let us look at the 
MF-diagram of [\textsc{g} 0.25–0.5], so that we can
see whether there are any other words in use for expressing
the same meaning (\figref{g0d2505.f}).
`fepi' has indeed emerged as dominant for this meaning, but 
this has not been without an intense struggle.
The tendency for synonymy to get damped is clearly present. 
Even though the lexicon occasionally destabilises, 
the lateral inhibition and the positive feedback loop between use 
and success causes self-organised MF-coherence. 
There is however something curious going on. 
In the early phases `xu' was dominant. Why did it 
destabilise and how has `fepi' has managed to become the
dominant expression for [\textsc{g} 0.25–0.5]?

>>>>

Wittgenstein proposed to study the use of language in terms of 
games. This notion captures that the most basic form of
language involves a social interaction between individuals
within a specific setting, which acts as a context restricting
the possible meanings. The interaction is played out 
along a conventionalised set of rules, as is chess or any
other game. The notion of a game captures
that a language interaction has a purpose, for example,
to identify an object in a context, to gather 
information, to transmit emotion, to try and
invoke an action by another person, etc. Words get \enlargethispage{2\baselineskip}
their meaning as part of a language game. Language is a tool 
fully integrated in the rest of human social activity.\footnote{See in particular \cite{Wittgenstein:1953}.}

@@

For the interpretation of an utterance, the main 
flow of information goes in the other direction. The
lexical layer sends a variety of hypotheses to 
the conceptual layer and the conceptual layer makes use of data 
from the perceptual layer to see which conceptualisation
yields a referent. Several solutions are considered 
by the conceptual layer because words are typically 
ambiguous and so the perceptual 
layer must produce the appropriate segmentations and 
sensory characteristics to test each solution.
Even if a sensory channel was not considered to be 
very salient by the hearer, it may have been 
used by the speaker in his conceptualisation of the 
scene and hence the hearer's perceptual 
processes must actively try to seek in the 
image the information required to see whether it 
is applicable to the specific
context. The interpretation of an utterance
thus strongly takes the real world context into account. 
The utterance literally influences the way the 
hearer sees the world. The two-way flow (from perception
to conceptualisation and language and from language
and conceptualisation to perception) resolves one of the 
paradoxes of meaning discussed in \chapref{chap:4}. The 
clear picture of reality that we consciously experience
results from a dynamical process in which local constraints 
keep propagating until a globally coherent solution
emerges.\footnote{Dynamical systems which enter into an attractor
based on a similar relaxation process have been 
widely studied, particularly in physics. One of the 
prototypical examples is a spin glass whose magnetic
states keep switching until a globally coherent
state is reached.}

@@

\begin{figure}
\begin{center}
\includegraphics[width=0.8\columnwidth]{chap7/figs/plate-10.pdf}
\end{center}
\caption{Three examples of segmented images. The 
topic is indicated by a dashed bounding box in the 
image of the speaker. Segments which are too small 
are ignored. The topics have all been conceptualised
as being `to the right' and so the same word 
\textit{gofubo} has been used to refer to them. }
\label{fig:plate-10}
\end{figure}

@@

The left image is that of the hearer {\bfshape a1}, the right one
that of the speaker, {\bfshape a2}.
{\bfshape a2} had already invented the word \textit{gofubo} for 
[\textsc{hpos} 0.5–1.0] (to the right) in an earlier game, 
but the word was not acquired in that game
by {\bfshape a1} because he still missed the 
appropriate distinction. Meanwhile the \textsc{hpos} category 
is available to {\bfshape a1} due to an
expansion of his discrimination networks
and so he can store the word: 
\begin{verbatim}
Game 3 
  a2 is the speaker. a1 is the hearer. 
  a2 segments the context into 2 objects: 
       object-0 object-1
  a2 chooses object-0 as the topic 
  a2 categorises the topic as [HPOS 0.5–1.0]
  a2 says: gofubo
  a1 does not know gofubo
  a1 says: gofubo?
  a2 points to object-0
  a1 categorises the topic as [HPOS 0.5–1.0]
  a1 stores gofubo as [HPOS 0.5–1.0]
\end{verbatim}
This game proceeds essentially like in the computer
simulations studied before. The major difference is
that now real images have been used, the objects
considered are the outcome of segmentation processes,
the sensory characteristics have been derived from the 
image itself, and the pointing has been done by 
physically moving the cameras.

@@

Immediately the agents apply this word to very 
different scenes, such as the ones in \figref{fig:plate-10} (middle and bottom). The middle picture
shows on the left the segmented scene
from the speaker in game 5, and on the right the one
from the hearer. In game 5, the figures are blue
rectangles. They are much further apart than 
the two circles in game 3. Nevertheless, 
after scaling, they are categorised and
conceptualised similarly and therefore
the same word could be used effectively. The concept 
of [\textsc{hpos} 0.5–1.0] and hence the word \textit{gofubo} 
is general {\itshape from the very start}. 
The agents do not need to see many examples because they do 
not use inductive generalisation. Instead, 
the \textsc{hpos} category is constructed in a top-down fashion 
as soon as the \textsc{hpos} channel has been salient
and is immediately available for use in the 
discrimination game and hence in verbalising 
the scene. This explains why the word-meaning acquisition
process observed in the experiments goes so 
amazingly fast.

The bottom of \figref{fig:plate-10}
shows yet another scene where the word \textit{gofubo} 
was used with success. The topic is the rightmost
shape in the scene and so [\textsc{hpos} 0.5–1.0] is once more 
distinctive. To an outside observer it may look 
like the agents have performed a gigantic inductive
leap, but this is not the case at all. The agents
do not try to abstract the commonalities from 
different examples but construct distinctions in a top 
down fashion and try to apply them to the perceived image.

@@

There is already a word (\textit{gofubo}) which has
a score of 1.0! After 100 games, the lexicon has become more 
solid and now looks as in \tabref{tab:goubo}.

@@

\begin{table}
\begin{center}
\begin{tabular}{lll}
\lsptoprule
{\itshape channel}& {\itshape obj-0} & {\itshape obj-1}\\ \midrule
\textsc{hpos} & 0.08 & 0.09\\ 
\textsc{vpos} & 0.21 & 0.10\\ 
\textsc{height} & 0.10 & 0.15\\ 
\textsc{width} & 0.42 & 0.11\\ 
\textsc{area} & 0.34 & 0.16\\ 
\textsc{r} & 0.33 & 0.31\\ 
\textsc{g} & 0.68 & 0.65\\ 
\textsc{b} & 0.52 & 0.50\\ 
\lspbottomrule
\end{tabular}
\caption{\label{tab:game126}Sensory data for game 126.}
\end{center}
\end{table}
Clearly width is the most salient channel for the speaker. 
It is therefore chosen and because the 
agents have already grown distinctions on this channel, 
discrimination succeeds and a new word can be
constructed and stored. 
\begin{verbatim}
Game 126 
  a2 is the speaker. a1 is the hearer. 
  a2 segments the context into 2 objects: 
       object-0 object-1
  a2 chooses object-0 as the topic 
  a2 categorises the topic as [WIDTH 0.5–1.0]
  a2 creates a new word: vaviwumu
  a2 says: vaviwumu
  a1 does not know vaviwumu
  a1 says: vaviwumu?
  a2 points to object-0
  a1 categorises the topic as [WIDTH 0.5–1.0]
  a1 stores vaviwumu as [WIDTH 0.5–1.0]
\end{verbatim}
\figref{fig:plate-13} (bottom) contains another example
of segmented images which have stimulated conceptual
growth and hence an expansion of the lexicon. 
In this case, categorisations on the \textsc{height} channel are
relevant and a word developed for tall.

@@

\begin{figure}
\begin{center}
\includegraphics[width=0.8\columnwidth]{chap7/figs/plate-15-vpos.pdf}
\end{center}
\caption{The image segments from the top game require 
distinctions on the \textsc{vpos} channel and 
caused the creation of a word for `upper'. The image 
segments in the middle led to the use of refined
distinctions on the \textsc{hpos} channel, so as to 
identify the middle square. The topic in the 
image segments at the bottom was done with a 
conjunction of categories: blue and tall.}
\label{fig:plate-15}
\end{figure}

@@

\begin{table}
\begin{center}
\begin{tabular}{lllll}
\lsptoprule
{\itshape Word}&{\itshape Meaning}&{\itshape Translation} & {\bfshape a1}&{\bfshape a2} \\ \midrule
wawosido & [\textsc{hpos} 0.0–0.5] &left&0.9&0.4\\ 
gixepo & [\textsc{hpos} 0.0–0.25] & very left&0.4&0.4\\ 
wonuxa & [\textsc{hpos} 0.0–0.25] & very left&0.10&0.0\\ 
meluri & [\textsc{hpos} 0.25–0.5] &medium left&0.4&0.4\\ 
gofubo & [\textsc{hpos} 0.5–1.0]& right&1.0&1.0\\ 
vokomutu & [\textsc{hpos} 0.5–0.75] &medium right&0.4&0.4\\ 
buwonipo & [\textsc{hpos} 0.75–0.875] &strongly right&0.1&0.0\\ 
wiwigapo & [\textsc{vpos} 0.0–0.5] &down&0.5&0.4\\ 
putuwenu & [\textsc{vpos} 0.5–1.0]&up & 0.5&0.5\\ 
vaviwumu & [\textsc{width} 0.5–1.0]&wide & 0.5&0.5\\ 
pesidumu & [\textsc{area} 0.0–0.5]&small& 0.2&0.2\\ 
fozumoba & [\textsc{area} 0.5–1.0]&large & 1.0&1.0\\ 
wefoto & [\textsc{r} 0.0–0.5]& low redness &0.2&0.1\\ 
togene & [\textsc{r} 0.5–1.0]& high redness &0.9&1.0\\ 
fumudanu & [\textsc{g} 0.0–0.5]& low greenness &0.7&0.9\\ 
puxedu & [\textsc{g} 0.5–1.0]& high greenness &0.9&0.9\\ 
\lspbottomrule
\end{tabular}
\caption{\label{tab:upper}Population lexicon after 200 games.}
\end{center}
\end{table}
Note that new words have come into the lexicon 
for \smplenquote{up} (\textit{putuwenu}), \smplenquote{down} (\textit{wiwigapo}), and 
\smplenquote{wide} (\textit{vaviwumu}). The discrimination trees of 
the \textsc{width} and \textsc{height} channel have started to 
expand (\figref{discri200} bottom).

@@

\begin{table}
\begin{center}
\begin{tabular}{llll}
\lsptoprule
{\itshape channel}& {\itshape obj-0} & {\itshape obj-1} & {\itshape obj-2}\\ 
\textsc{hpos} & 0.03 & 0.15 & 0.33\\ 
\textsc{vpos} & 0.29 & 0.29 & 0.01\\ 
\textsc{height} & 0.37 & 0.39 & 0.07\\ 
\textsc{width} & 0.10 & 0.08 & 0.27\\ 
\textsc{area} & 0.29 & 0.22 & 0.16\\ 
R & 0.95 & 0.97 & 0.97 \\ 
G & 0.33 & 0.40 & 0.93\\ 
B & 0.35 & 0.44 & 0.26\\ 
\lspbottomrule
\end{tabular}
\caption{\label{tab:game128}Sensory data for game 128.}
\end{center}
\end{table}
Context-scaling further
amplifies this difference which shows that it is not 
always beneficial to do so. {\bfshape a1} uses
the horizontal axis, creating a new word \textit{bagaxe} 
meaning `very left' [\textsc{hpos} 0.0–0.25] whereas {\bfshape a2}, for
whom the area is the most salient characteristic, 
associates this word with [\textsc{area} 0.5–1.0] (large). 
This example shows that grounding introduces 
additional risks for the introduction of semantic
incoherence in a group's lexicon.

Due to perceptual anomalies, a game may fail even though 
both agents already have the same meaning for the 
same word. This happens when this meaning yields different 
objects (or no objects at all) for the speaker and the hearer. 
As a consequence, the hearer adopts another meaning for 
the word which starts to compete with the one he already had. 
This happened in the following game (game 127)
based on the segmented images shown in \figref{fig:plate-16} (bottom). The speaker (left image) has identified
only two objects, but the hearer three. The 
speaker (left image) uses green as distinguishing 
characteristic which is indeed appropriate. but 
because the hearer's third object (the top rectangle) is 
also green, this distinction fails for him.  
Even though the hearer had already a well established
meaning for \textit{puxedu}, he associates a new meaning 
based on the \textsc{width}-channel with this word because this 
channel is now the most salient. 
\begin{verbatim}
Game 127
  a1 is the speaker. a2 is the hearer. 
  a1 segments the context into 2 objects: 
       s-object-0 s-object-1
  a1 chooses object-0 as the topic 
  a1 considers as salient G R AREA HPOS 
  a1 categorises the topic as 
       [HPOS 0.0–0.5] [HPOS 0.0–0.125]
       [AREA 0.0–0.5] [R 0.0–0.5] [G 0.5–1.0]
  a1 has the words
       puxedu for [G 0.5–1.0] (1.0)
       wawosido for [HPOS 0.0–0.5] (0.7)
       pesidimu for [AREA 0.0–0.5] (0.2)
       wefoto for [R 0.0–0.5] (0.2)
       buwonipo for [G 0.5–1.0] (0.0)
  a1 says: puxedu
  a2 segments the context into 3 objects: 
       h-object-0 h-object-1 h-object-2
  a2 interprets puxedu as
       [G 0.5–1.0] (0.20)
  a2 identifies h-object-0 h-object-1
  a2 says: puxedu?
  a1 points to s-object-0
  a2 categorises the topic as [WIDTH 0.5–1.0]
  a2 stores puxedu as [WIDTH 0.5–1.0]
\end{verbatim}

@@

The partial landscape in \figref{RMF1} (taken from
an actual experiment) contains an example where the agents use
two possible meanings for conceptualising {\itshape situation2}, namely
[\textsc{gray} 0.0–0.25] (very light) and [\textsc{hpos} 0.5–0.75]
(medium right ). The words \textit{katapu} and \textit{tisame} lexicalise
[\textsc{gray} 0.0–0.25] and \textit{wobo} and \textit{tisame} [\textsc{hpos} 0.5–0.75]. 
Each meaning has therefore two synonyms and \textit{tisame} is 
ambiguous; it can mean both [\textsc{gray} 0.0–0.25]
and [\textsc{hpos} 0.5–0.75].
\enlargethispage{1em}
Three words
are used to communicate the situation: `katapu', `tisame', 
and `wobo'. Such a structure is typical for grounded
lexicon evolution and complexity rapidly increases when the
same meanings are used to communicate about other
situations (which is obviously very common and indeed desirable).

@@

The RF-diagram\is{RF-diagram} in \figref{RF-diagram} shows the 
frequency with which certain words were used to communicate 
a particular situation. We see that in
the beginning the word `desepu' has been dominant, then 
there is a period of turbulence in which different 
words compete, but after a while a new word \textit{va} wins 
the competition and becomes the dominant way to communicate 
about this situation. \figref{FM-diagram} shows another\is{FM-diagram}
competition diagram plotting the evolution of the FM-relation, for 
the word form \textit{va}, in other words the frequencies of 
all the meanings that co-occurred with the word \textit{va}. 
We see an early peak when \textit{va} was used in 70 \% of the 
games with the meaning [\textsc{b} 0.3125–0.375], i.e. a particular
shade of blue. Then there is a struggle during which additional
distinctions (on the \textsc{red} and \textsc{vpos}-channels) are competing
for the dominant meaning of \textit{va}. 
[\textsc{r} 0.0–0.125] finally becomes dominant.

\begin{figure}[htbp]
  \centerline{\includegraphics[width=.80\textwidth]{chap7/figs/fm.pdf}}
\caption{\label{FM-diagram}This FM-diagram shows
the frequencies of each form-meaning pair with 
the form equal to \textit{va} in a series of 5000 games.
A disambiguating situation occurs in game 3000
causing the loss of one meaning of \textit{va}.}
\end{figure}

These competition diagrams are an important tool to try 
and make sense of the ontological and lexical evolution
taking place in evolving groups of agents as they are
playing their language games. Typically we pick
a dominating word, for example the word \textit{va} in 
\figref{RF-diagram}, and try to understand
why it has become dominant. 
The FM-diagram for \textit{va} (\figref{FM-diagram}) explains
part of the story. Three stable meanings for \textit{va} have emerged
at  around 1000 games: 
[\textsc{r} 0.0–0.125], [\textsc{b} 0.3125–0.375], and [\textsc{vpos} 0.25–0.5]. 
They are all equally adequate for distinguishing the object
\textit{va} designates, and there are no situations yet
that would have forced the disambiguation of \textit{va}. 
In game 3000, the environment (which is continuously 
changing in this experiment) produces a scene in which
a category which was distinctive for the object designated by 
\textit{va} is no longer distinctive. The lexicon adapts 
immediately. Around game 3000 the \textsc{vpos}-based meaning disappears, 
and the distinction based on \textsc{red} shoots up and becomes dominant.

@@

\subsection{The story of \textit{fepi}}

\begin{figure}[htbp]
  \centerline{\includegraphics[width=.80\textwidth]{chap7/figs/FR-FEPI.pdf}}
\caption{\label{fr-fepi}FR-diagram showing the frequencies
of the objects referred to by the word
\textit{fepi}. \textit{fepi} is consistently used for \emph{O3} as well as
\emph{O5}.}
\end{figure}
Only looking at macroscopic measures like communicative
success hides away the
interesting rich lexical dynamics that unfolds
in the population. Let me examine just one word, \textit{fepi}. 
By looking at the FR-diagram, we see that this word is 
used consistently for identifying two objects
(\emph{O3} and \emph{O5}) in a certain set of contexts
(\figref{fr-fepi}).

\begin{figure}[htbp]
  \centerline{\includegraphics[width=.80\textwidth]{chap7/figs/FM-FEPI.pdf}}
\caption{\label{fm-fepi}FM-diagram showing the different
meanings of \textit{fepi}. One meaning [\textsc{g} 0.25–0.5] dominates.}
\end{figure}
We can see the meanings of \textit{fepi}, by 
inspecting its FM-diagram (\figref{fm-fepi}). 
The dominant meaning of \textit{fepi} is 
a particular shade of green [\textsc{g} 0.25–0.5]. There are 
some other competing meanings (including [\textsc{gray} 0.25–0.3125])
but most of them are hardly ever used. We observe clearly 
the tendency for ambiguity to get damped.

What about synonymy? Let us look at the 
MF-diagram of [\textsc{g} 0.25–0.5], so that we can
see whether there are any other words in use for expressing
the same meaning (\figref{g0d2505.f}).
\textit{fepi} has indeed emerged as dominant for this meaning, but 
this has not been without an intense struggle.
The tendency for synonymy to get damped is clearly present. 
Even though the lexicon occasionally destabilises, 
the lateral inhibition and the positive feedback loop between use 
and success causes self-organised MF-coherence. 
There is however something curious going on. 
In the early phases \textit{xu} was dominant. Why did it 
destabilise and how has \textit{fepi} has managed to become the
dominant expression for [\textsc{g} 0.25–0.5]?
